---
title: "Stat 210 Final Project"
Author: "Brian Bieza"
Date: 4-29-2026
format: pdf
---

## Load Libraries

```{r message=FALSE, warning=FALSE}
library(ggrepel)
library(ggpmisc)
library(tidyverse)
library(vegan)
```

## Loading in CSV's

```{r}
Fish <- read.csv("Group_3_Mastersheet.csv")
```

## Cleaning Data

```{r}
Fish_Cleaned <- Fish %>%
  select(Location, Species, Observed_Behavior, Adjacent_Habitat) %>% 
  mutate(Location = recode(Location, "McAbee" = "Mcabee"),
         Species = recode(Species, "Black_Eyed_Goby" = "Black_Eyed_Goby"))

# added Northern California/Southern California
Fish_Cleaned_2 <- Fish_Cleaned %>%
  filter(!str_detect(Species, "YOY|UFO")) %>% 
  mutate(Region = case_when(
    Location %in% c("NMON", "SMON", "Carmel_Beach") ~ "Carmel_Bay",
    Location %in% c("BW_Wall", "BW_Reef", "Mcabee")  ~ "Monterey_Bay",
    Location == "Point_Pinos"                         ~ "Transition_Zone",
    TRUE ~ NA_character_),
  Species = recode(Species, "Black_Eyed_Goby" = "Black_Eyed_Goby")
  )
```

## Diversity by Site

```{r}
# Shannon diversity by individual site

# Build a species x site abundance matrix
species_matrix <- Fish_Cleaned_2 %>%
  group_by(Location, Species) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Species, values_from = count, values_fill = 0) %>%
  column_to_rownames("Location")

# Calculate Shannon H' per site
shannon <- diversity(species_matrix, index = "shannon")

shannon_df <- shannon %>%
  as.data.frame() %>%
  rownames_to_column("Location") %>%
  rename(Shannon = ".") %>%
  left_join(
    Fish_Cleaned_2 %>% distinct(Location, Region),
    by = "Location"
  )


print(shannon_df)
```

## Diversity by Region

```{r}
# Build species x region abundance matrix
Region_Matrix <- Fish_Cleaned_2 %>%
  group_by(Region, Species) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Species, values_from = count, values_fill = 0) %>%
  column_to_rownames("Region")

# Calculate Shannon diversity per region
shannon_region <- diversity(Region_Matrix, index = "shannon")

# Convert to tidy dataframe
shannon_df <- shannon_df %>%
  mutate(Location = factor(Location, levels = c(
    "BW_Wall", "BW_Reef", "Mcabee",      # Monterey Bay
    "Point_Pinos",                         # Transition Zone
    "NMON", "SMON", "Carmel_Beach"        # Carmel Bay
  )),
  Region = factor(Region, levels = c("Monterey_Bay", "Transition_Zone", "Carmel_Bay")))

# View results
print(shannon_df)
```
## Shannon Diversity by Site

```{r}
ggplot(shannon_df, aes(x = Location, y = Shannon, fill = Region)) +
  geom_col(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(Shannon, 2)), hjust = -0.2, size = 3.5) +
  scale_fill_manual(values = c(
    "Carmel_Bay"      = "steelblue",
    "Monterey_Bay"    = "orange",
    "Transition_Zone" = "red"
  )) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right") +
  expand_limits(y = max(shannon_df$Shannon) * 1.15) +
  labs(
    title = "Shannon Diversity by Site",
    x = NULL,
    y = "Shannon H'",
    fill = "Region"
  )
```

## Species Richness

```{r}
# Richness per site
richness_site <- Fish_Cleaned_2 %>%
  group_by(Location, Region) %>%
  summarise(Richness = n_distinct(Species), .groups = "drop")

# Richness per region
richness_region <- Fish_Cleaned_2 %>%
  group_by(Region) %>%
  summarise(Richness = n_distinct(Species), .groups = "drop")

# View results
print(richness_site)
print(richness_region)
```

## Diversity Dataframe

```{r}
diversity_summary <- shannon_df %>%
  left_join(richness_site, by = c("Location", "Region"))

print(diversity_summary)
```

## Data Cleaning For Behavior Associations

```{r}
# Remove YOY, UFO, and any blank/NA entries
Fish_Cleaned_3 <- Fish_Cleaned_2 %>%
  filter(!str_detect(Species, "YOY|UFO")) %>%
  filter(Species != "" & !is.na(Species)) %>%
  filter(Observed_Behavior != "" & !is.na(Observed_Behavior)) %>%
  mutate(Species = recode(Species, "Black_Eyed_Gobeye" = "Black_Eyed_Goby"))

# Check what you're working with
unique(Fish_Cleaned_3$Species)
unique(Fish_Cleaned_3$Observed_Behavior)
```

## Behavior Associations

```{r}
# Behavior x Species
behavior_species <- Fish_Cleaned_3 %>%
  count(Species, Observed_Behavior) %>%
  pivot_wider(names_from = Observed_Behavior, values_from = n, values_fill = 0) %>%
  column_to_rownames("Species")

chisq_species <- chisq.test(behavior_species)
print(chisq_species)
fisher_species <- fisher.test(behavior_species, simulate.p.value = TRUE, B = 10000)
print(fisher_species)
```

# Behavior x Location

```{r}
behavior_location <- Fish_Cleaned_3 %>%
  count(Location, Observed_Behavior) %>%
  pivot_wider(names_from = Observed_Behavior, values_from = n, values_fill = 0) %>%
  column_to_rownames("Location")

chisq_location <- chisq.test(behavior_location)
fisher_location <- fisher.test(behavior_location, simulate.p.value = TRUE, B = 10000)
```

## Species x Behavior Heatmap

```{r}
Species_Behavior_Heatmap <- chisq_species$residuals %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  pivot_longer(-Species, names_to = "Observed_Behavior", values_to = "Residual") %>%
   mutate(
    Species = str_replace_all(Species, "_", " "),
    Observed_Behavior = str_replace_all(Observed_Behavior, "_", " ")
  ) %>%
  ggplot(aes(x = Observed_Behavior, y = Species, fill = Residual)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "coral", midpoint = 0) +
  theme_minimal() +
  labs(title = "Species x Behavior Association",
       x = "Observed Behavior", y = "Species")
print(Species_Behavior_Heatmap)
ggsave("Species_Behavior_Heatmap.pdf", plot = Species_Behavior_Heatmap, width = 10, height = 8, dpi = 300)
```

## Location x Behavior Heatmap

```{r}
Location_Behavior_Heatmap <- chisq_location$residuals %>%
  as.data.frame() %>%
  rownames_to_column("Location") %>%
  pivot_longer(-Location, names_to = "Observed_Behavior", values_to = "Residual") %>%
  left_join(
    Fish_Cleaned_3 %>% distinct(Location, Region),
    by = "Location"
  ) %>%
  mutate(
    Location = str_replace_all(Location, "_", " "),
    Observed_Behavior = str_replace_all(Observed_Behavior, "_", " "),
    Region = str_replace_all(Region, "_", " ")
  ) %>%
  ggplot(aes(x = Observed_Behavior, y = Location, fill = Residual)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "orange", midpoint = 0) +
  facet_grid(Region ~ ., scales = "free_y", space = "free_y") +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "gray90", color = NA),
    strip.text = element_text(face = "bold", size = 10),
    panel.spacing = unit(0.5, "lines")
  ) +
  labs(title = "Location x Behavior Association",
       x = "Observed Behavior", y = "Location")
print(Location_Behavior_Heatmap)
ggsave("Location_Behavior_Heatmap.pdf", plot = Location_Behavior_Heatmap, width = 10, height = 8, dpi = 300)
```

## Adjacent Habitat x Species

```{r}
habitat_species <- Fish_Cleaned_3 %>%
  count(Species, Adjacent_Habitat) %>%
  pivot_wider(names_from = Adjacent_Habitat, values_from = n, values_fill = 0) %>%
  column_to_rownames("Species")

fisher_habitat_species <- fisher.test(habitat_species, simulate.p.value = TRUE, B = 10000)
print(fisher_habitat_species)
```

## Adjacent Habitat x Location

```{r}
habitat_location <- Fish_Cleaned_3 %>%
  count(Location, Adjacent_Habitat) %>%
  pivot_wider(names_from = Adjacent_Habitat, values_from = n, values_fill = 0) %>%
  column_to_rownames("Location")

fisher_habitat_location <- fisher.test(habitat_location, simulate.p.value = TRUE, B = 10000)
print(fisher_habitat_location)
chisq_habitat_species <- chisq.test(habitat_species)
```

## Habitat x Species heatmap

```{r}
Habitat_Species_Heatmap <- chisq_habitat_species$residuals %>%
  as.data.frame() %>%
  rownames_to_column("Species") %>%
  pivot_longer(-Species, names_to = "Adjacent_Habitat", values_to = "Residual") %>%
  mutate(
    Species = str_replace_all(Species, "_", " "),
    Adjacent_Habitat = str_replace_all(Adjacent_Habitat, "_", " ")
  ) %>%
  ggplot(aes(x = Adjacent_Habitat, y = Species, fill = Residual)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "coral", midpoint = 0) +
  theme_minimal() +
  labs(title = "Species x Adjacent Habitat Association",
       x = "Adjacent Habitat", y = "Species")
print(Habitat_Species_Heatmap)
ggsave("Habitat_Species_Heatmap.pdf", plot = Habitat_Species_Heatmap, width = 10, height = 8, dpi = 300)
```

## Habitat x Location heatmap

```{r}
chisq_habitat_location <- chisq.test(habitat_location)

Habitat_Location_Heatmap <- chisq_habitat_location$residuals %>%
  as.data.frame() %>%
  rownames_to_column("Location") %>%
  pivot_longer(-Location, names_to = "Adjacent_Habitat", values_to = "Residual") %>%
  left_join(Fish_Cleaned_3 %>% distinct(Location, Region), by = "Location") %>%
  mutate(
    Location = str_replace_all(Location, "_", " "),
    Adjacent_Habitat = str_replace_all(Adjacent_Habitat, "_", " "),
    Region = str_replace_all(Region, "_", " ")
  ) %>%
  ggplot(aes(x = Adjacent_Habitat, y = Location, fill = Residual)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "coral", midpoint = 0) +
  facet_grid(Region ~ ., scales = "free_y", space = "free_y") +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "gray90", color = NA),
    strip.text = element_text(face = "bold", size = 10),
    panel.spacing = unit(0.5, "lines")
  ) +
  labs(title = "Location x Adjacent Habitat Association",
       x = "Adjacent Habitat", y = "Location")
print(Habitat_Location_Heatmap)
ggsave("Habitat_Location_Heatmap.pdf", plot = Habitat_Location_Heatmap, width = 10, height = 8, dpi = 300)
```

## Species by Site Heatmap

```{r}
site_species <- Fish_Cleaned_2 %>%
  filter(!is.na(Region), !Species %in% c("UFO", "YOY", "OYT_YOY", "KGB_YOY")) %>%
  group_by(Region, Location, Species) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    Species = recode(Species, "Black_Eyed_Gobeye" = "Black_Eyed_Goby"),
    Species = str_replace_all(Species, "_", " "),
    Location = str_replace_all(Location, "_", " "),
    Region = str_replace_all(Region, "_", " ")
  )

site_species <- site_species %>%
  mutate(Location = factor(Location, levels = c(
    "NMON", "SMON", "Carmel Beach",    # Carmel Bay
    "BW Wall", "BW Reef", "Mcabee",    # Monterey Bay
    "Point Pinos"                       # Transition Zone
  )))

Species_Presence_Heatmap <- ggplot(site_species, aes(x = Location, y = Species, fill = count)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue",
                      name = "Observations") +
  facet_grid(~ Region, scales = "free_x", space = "free_x") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    axis.text.y      = element_text(size = 9),
    strip.text       = element_text(face = "bold", color = "white"),
    strip.background = element_rect(fill = c("steelblue", "darkgreen", "orange"), color = NA),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  ) +
  labs(x = NULL, y = NULL,
       title = "Species presence by site",
       subtitle = "Grouped by biogeographic region")
print(Species_Presence_Heatmap)
ggsave("Species_Presence_Heatmap.pdf", plot = Species_Presence_Heatmap, width = 10, height = 8, dpi = 300)
```

## Density of Fish per site

```{r}
transect_area_m2 <- 120 

density_df <- Fish_Cleaned_3 %>%
  group_by(Location, Region) %>%
  summarise(total_fish = n(), .groups = "drop") %>%
  mutate(
    density = total_fish / transect_area_m2,
    Location = str_replace_all(Location, "_", " "),
    Region = str_replace_all(Region, "_", " ")
  ) %>%
  mutate(
    Location = factor(Location, levels = c(
      "BW Wall", "BW Reef", "Mcabee",          # Monterey Bay
      "Point Pinos",                             # Transition Zone
      "SMON", "NMON", "Carmel Beach"            # Carmel Bay
    )),
    Region = factor(Region, levels = c("Monterey Bay", "Transition Zone", "Carmel Bay"))
  )
print(density_df)
```

# Density By Site

```{r}
Density_Heatmap <- ggplot(density_df, aes(x = Location, y = density, fill = Region)) +
  geom_col(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(density, 3)), 
            hjust = -0.2, size = 3.5, color = "gray30") +
  coord_flip() +
  scale_fill_manual(values = c(
    "Carmel Bay"      = "steelblue",
    "Monterey Bay"    = "orange",
    "Transition Zone" = "red"
  )) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank()
  ) +
  labs(
    title = "Fish density by site",
    subtitle = paste("Per", transect_area_m2, "m² transect area"),
    x = NULL,
    y = "Fish density (individuals / m²)"
  )
print(Density_Heatmap)
ggsave("Density_Heatmap.pdf", plot = Density_Heatmap, width = 10, height = 8, dpi = 300)
```

## Define SoCal VS. NorCal

```{r}
unique(Fish_Cleaned_2$Species)
# Define SoCal and NorCal species using exact names from dataset
socal_species <- c("Senorita", "CA_Sheephead_F", "Halfmoon", 
                   "Opaleye", "Kelp_Bass",                            "Blacksmith")

norcal_species <- c("Blue_Rockfish", "Pile_Perch", "Olive_Rockfish", "Kelp_Perch",
                    "Painted_Greenling", "Yellowtail_Rockfish", "Kelp_Greenling",
                    "Rubberlip_Perch", "Black_Eyed_Goby", "Lingcod", 
                    "Striped_Perch", "Copper_Rockfish", "Bocaccio", "Black_Perch", "Kelp_Rockfish")

# Filter out ambiguous IDs and classify
Fish_Classified <- Fish_Cleaned_3 %>%
  filter(!Species %in% c("YOY", "Rockfish", "Perch", "KGB_YOY", "OYT_YOY", "UFO")) %>%
  mutate(Bioregion = case_when(
    Species %in% socal_species  ~ "Southern CA",
    Species %in% norcal_species ~ "Northern CA",
    TRUE ~ NA_character_
  ))
```

### NMDS Plot

# Build site x species matrix from classified fish

```{r}
nmds_matrix <- Fish_Classified %>%
  group_by(Location, Species) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = Species, values_from = count, values_fill = 0) %>%
  column_to_rownames("Location")
```

# Run NMDS

```{r}
set.seed(123)
nmds <- metaMDS(nmds_matrix, distance = "bray", k = 2, trymax = 100)
print(nmds)  # check stress score
```

# Extract site scores and join Region + SoCal proportion

```{r}
site_scores <- as.data.frame(scores(nmds, display = "sites")) %>%
  rownames_to_column("Location") %>%
  left_join(Fish_Classified %>% distinct(Location, Region), by = "Location") %>%
  left_join(
    Fish_Classified %>%
      group_by(Location) %>%
      summarise(
        prop_socal = mean(Bioregion == "Southern CA"),
        .groups = "drop"
      ),
    by = "Location"
  ) %>%
  mutate(
    Location = str_replace_all(Location, "_", " "),
    Region = str_replace_all(Region, "_", " ")
  )
```

# Extract species scores

```{r}
species_scores <- as.data.frame(scores(nmds, display = "species")) %>%
  rownames_to_column("Species") %>%
  left_join(
    data.frame(
      Species = c(socal_species, norcal_species),
      Bioregion = c(rep("Southern CA", length(socal_species)),
                    rep("Northern CA", length(norcal_species)))
    ),
    by = "Species"
  ) %>%
  mutate(Species = str_replace_all(Species, "_", " "))
```

# Fit SoCal proportion as environmental vector

```{r}
socal_vector <- Fish_Classified %>%
  group_by(Location) %>%
  summarise(prop_socal = mean(Bioregion == "Southern CA"), .groups = "drop") %>%
  column_to_rownames("Location")

env_fit <- envfit(nmds, socal_vector, permutations = 999, na.rm = TRUE)
print(env_fit)  # check significance of vector
```

# Extract vector coordinates for plotting

```{r}
vector_df <- as.data.frame(scores(env_fit, display = "vectors")) %>%
  rownames_to_column("Variable") %>%
  mutate(Variable = "SoCal Proportion")
```

# NMDS Plot

```{r}
nmds_plot <- ggplot() +
  geom_point(data = site_scores,
             aes(x = NMDS1, y = NMDS2, fill = Region, size = prop_socal),
             shape = 21, color = "white", stroke = 0.5) +
  geom_text_repel(data = site_scores,
                  aes(x = NMDS1, y = NMDS2, label = Location),
                  size = 3,fontface = "bold", color = "gray30", max.overlaps = 20,
                  box.padding = 0.5) +
  geom_segment(data = species_scores,
               aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2, color = Bioregion),
               arrow = arrow(length = unit(0.2, "cm")), linewidth = 0.5, alpha = 0.6) +
  geom_segment(data = vector_df,
               aes(x = 0, y = 0, xend = NMDS1 * 0.8, yend = NMDS2 * 0.8),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black", linewidth = 1, alpha = 0.4) +
  geom_text(data = vector_df,
            aes(x = NMDS1 * 0.85, y = NMDS2 * 0.85, label = Variable),
            size = 3.5, fontface = "bold", hjust = -0.1) +
  geom_text_repel(data = species_scores,
                aes(x = NMDS1, y = NMDS2, label = Species, color = Bioregion),
                size = 3.5, max.overlaps = 20,
                box.padding = 1.75, segment.alpha = 0.3,
                segment.color = "black", 
                force = 3,
                bg.color = "darkgrey", bg.r = 0.01,
                show.legend = FALSE) +
  scale_fill_manual(values = c(
    "Carmel Bay"      = "steelblue",
    "Monterey Bay"    = "darkgreen",
    "Transition Zone" = "red"
  )) +
  scale_color_manual(values = c(
    "Southern CA" = "orange",
    "Northern CA" = "steelblue"
  )) +
  scale_size_continuous(name = "SoCal proportion", range = c(3, 10)) +
  theme_minimal(base_size = 12) +
  guides(
  fill = guide_legend(override.aes = list(size = 6)),
  size = guide_legend(title = "SoCal proportion"),
  color = guide_legend(override.aes = list(size = 4))
) +
  theme(legend.position = "right") +
  labs(
    title = "NMDS of fish community composition",
    subtitle = "Arrow = SoCal proportion gradient | Point size = SoCal proportion | Species colored by bioregion",
    x = "NMDS1", y = "NMDS2",
    fill = "Region", color = "Bioregion"
  )
print(nmds_plot)
ggsave("nmds_plot.pdf", plot = nmds_plot, width = 10, height = 8, dpi = 300)
```

## Building proportion data

```{r}
diverging_df <- Fish_Classified %>%
  group_by(Location, Region) %>%
  summarise(
    prop_socal = mean(Bioregion == "Southern CA"),
    prop_norcal = mean(Bioregion == "Northern CA"),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(prop_socal, prop_norcal),
               names_to = "Bioregion",
               values_to = "Proportion") %>%
  mutate(
    Proportion = ifelse(Bioregion == "prop_norcal", -Proportion, Proportion),
    Bioregion = recode(Bioregion,
                       "prop_socal" = "Southern CA",
                       "prop_norcal" = "Northern CA"),
    Location = str_replace_all(Location, "_", " "),
    Region = str_replace_all(Region, "_", " "),
    Location = factor(Location, levels = c(
      "NMON", "SMON", "Carmel Beach",
      "BW Wall", "BW Reef", "Mcabee",
      "Point Pinos"
    ))
  )
```

## Diverging Bar Chart

```{r}
Socal_Norcal_Barchart <- ggplot(diverging_df, aes(x = Location, y = Proportion, fill = Bioregion)) +
  geom_col(color = "white", linewidth = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.8, color = "gray20") +
  scale_fill_manual(values = c(
    "Southern CA" = "orange",
    "Northern CA" = "steelblue"
  )) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x) * 100, "%"),
    limits = c(-1, 1)
  ) +
  facet_grid(~ Region, scales = "free_x", space = "free_x") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "gray90", color = NA),
    strip.text = element_text(face = "bold", size = 10),
    panel.spacing = unit(0.5, "lines"),
    legend.position = "bottom",
    panel.grid.major.y = element_blank()
  ) +
  labs(
    title = "Biogeographic composition by site",
    subtitle = "Proportion of Southern CA vs Northern CA species",
    x = NULL,
    y = "Proportion",
    fill = "Bioregion"
  )
print(Socal_Norcal_Barchart)
ggsave("Socal_Norcal_Barchart.pdf", plot = Socal_Norcal_Barchart, width = 10, height = 8, dpi = 300)
```
